\documentclass[journal]{IEEEtran}
\usepackage{amsmath,amsfonts}

\begin{document}

\title{paGating: A Parameterized Activation Gating Framework for Flexible and Efficient Neural Networks for GenAI}

% REAL AUTHOR INFORMATION FOR TITLE PAGE SUBMISSION
\author{Aaryan Guglani\\
        Department of Computer Science and Engineering\\
        RV College of Engineering\\
        Bengaluru, 560059 India\\
        Email: aaryanguglani.cs21@rvce.edu.in\\
        ORCID: 0009-0001-1086-1387\\[1em]
        Rajashree Shettar\\
        Professor, Department of Computer Science and Engineering\\
        RV College of Engineering\\
        Bengaluru, 560059 India\\
        Email: rajashreeshettar@rvce.edu.in\\
        ORCID: 0000-0001-6651-1768
}

\maketitle

\begin{abstract}
Activation gating has emerged as a powerful mechanism for modulating information flow within neural networks. This work introduces paGating, an open-source framework that unifies a family of parameterized activation gating units. Each unit exposes a tunable (or learnable) scalar $\alpha$ that balances classical gating activations—such as sigmoid, tanh, GELU, and Mish—with their un-gated counterparts. This simple yet expressive parameterization enables practitioners to interpolate continuously between standard feed-forward layers and fully gated alternatives without architectural surgery. Through systematic evaluation on GPT-2 Small with WikiText-103, the results demonstrate that paGating achieves consistent improvements in evaluation loss with validated results showing 1.9\% improvement when comparing parameterized gating ($\alpha=0.5$) against identity mapping ($\alpha=0.0$) while introducing zero parameter overhead.
\end{abstract}

\begin{IEEEkeywords}
Activation functions, Gating mechanisms, Parameterized neural networks, Transformers, Language modeling.
\end{IEEEkeywords}

% Contact Information Section
\section*{Author Contact Information}
\textbf{Corresponding Author:} Aaryan Guglani\\
\textbf{Email:} aaryanguglani.cs21@rvce.edu.in\\
\textbf{Address:} Department of Computer Science and Engineering, RV College of Engineering, Bengaluru, 560059 India\\
\textbf{ORCID:} 0009-0001-1086-1387

\textbf{Co-Author:} Rajashree Shettar\\
\textbf{Email:} rajashreeshettar@rvce.edu.in\\
\textbf{Address:} Department of Computer Science and Engineering, RV College of Engineering, Bengaluru, 560059 India\\
\textbf{ORCID:} 0000-0001-6651-1768

% Acknowledgments section with real funding/support information
\section*{Acknowledgments}
The authors acknowledge the use of AI-assisted tools including Claude (Anthropic) for manuscript preparation and code documentation. This research was conducted independently without external funding. The authors thank the open-source community for PyTorch, Transformers, and related libraries that enabled this work.

\end{document} 